---
title: "HollisT_Week4"
author: "Tim Hollis"
date: "`r Sys.Date()`"
output: html_document
editor_options: 
  markdown: 
    wrap: 72
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo=TRUE,warning=FALSE,message=FALSE)

#Libraries needed
library(dplyr)
library(tidyr)
library(ggplot2)
library(caret)
library(lubridate)
library(mice)
library(e1071)

games<-read.csv('vgchartz-2024.csv')
```

### Initial Data Review

```{r initial-data-review}

# Initial Data Review and Structure of dataset

# Dataset dimensions and column names
cat('Dataset Dimensions:',dim(games),'\n')
cat('Column Names:',names(games),'\n')

# Preview of first few rows
knitr::kable(head(games),caption='Preview of First Few Rows')

# Structure of dataset: data types and sample values
cat('\nStructure of dataset:\n')
str(games)

# Missing values summary: NA and empty strings
missing_summary<- data.frame(
  Column=names(games),
  NA_Count=colSums(is.na(games)),
  Empty_String_Count=colSums(games=='')
)
knitr::kable(missing_summary,caption='Missing Value Summary by Column')

# Diagnostic: flag columns with any missing values (NA or empty string)
cat('\nColumns with missing values (diagnostic only):\n')
print(names(games)[colSums(is.na(games))+colSums(games=='')>0])
```

### Initial data exploration

The dataset contains 64,016 rows and 14 columns, representing a
substantial archive of video game records. Column names span metadata
(title, console, genre, publisher, developer), performance metrics
(critic_score, total_sales, and regional sales breakdowns), and temporal
markers (release_date, last_update). The img column stores file paths to
box art images, which may be useful for filtering or visual reference.

A preview of the first few rows includes familiar titles such as Grand
Theft Auto V and Call of Duty: Black Ops 3, spanning platforms like PS2,
PS3, PS4, and Xbox 360. Sales figures are broken down by region
(na_sales, jp_sales, pal_sales, other_sales), and critic scores are
included for most entries, though some are missing.

The structure summary confirms that most columns are character or
numeric types. Notably, several numeric columns—including critic_score,
total_sales, and all regional sales figures—contain a high volume of NA
values. Additionally, some character columns such as developer,
release_date, and last_update include missing values stored as empty
strings ("") rather than NA.

A diagnostic check flagged multiple columns with missing data. However,
per assignment instructions, only the release_date column will be
imputed using data simulation. Other missing values will remain
untouched for this analysis.

```{r data-transformation}

# Create a copy for transformation
games_transformed<-games

# Convert release_date to Date format and identify missing values
games_transformed$release_date<-as.Date(games_transformed$release_date)
missing_dates<-is.na(games_transformed$release_date)
cat('Missing release dates:',sum(missing_dates),'\n')

# Convert character columns to appropriate types
games_transformed$console<-as.factor(games_transformed$console)
games_transformed$genre<-as.factor(games_transformed$genre)
games_transformed$publisher<-as.factor(games_transformed$publisher)
games_transformed$developer<-as.factor(games_transformed$developer)
```

### Preparation of data for analysis

To begin the transformation process, a working copy of the dataset was
created (games_transformed) to preserve the integrity of the original
data. The release_date column was converted to Date format, revealing
7,051 missing entries, which will be addressed later using data
simulation as specified in the assignment.

Several character columns—console, genre, publisher, and developer—were
converted to factors to ensure proper handling during modeling and
transformation. This step helps distinguish categorical variables from
numeric ones and prevents accidental scaling or normalization of
non-numeric data.

A subset of columns containing numeric values was identified for
transformation: critic_score, total_sales, na_sales, jp_sales,
pal_sales, and other_sales. These variables will undergo scaling, skew
correction, and normalization in the following steps to prepare the data
for analysis.

```{r scale-numeric-data}

# Define numerical columns to scale
numerical_cols <- c('critic_score','total_sales','na_sales','jp_sales','pal_sales','other_sales')

# Apply z-score standardization
scaled_data <- scale(games_transformed[numerical_cols])

# Assign scaled columns back to the dataset with "_scaled" suffix
for (i in seq_along(numerical_cols)) {
  col_name <- paste0(numerical_cols[i], "_scaled")
  games_transformed[[col_name]] <- scaled_data[, i]
}

# Preview scaled values
scaled_preview <- games_transformed[paste0(numerical_cols, "_scaled")]
knitr::kable(head(scaled_preview), caption = 'Scaled Numerical Columns')
```

### Results of Scaling:

To prepare the dataset for analysis, all selected numerical columns were
standardized using z-score scaling. This transformation centers each
variable around a mean of 0 and scales it to unit variance, allowing for
meaningful comparison across features with different units and ranges.

The scaled columns include:

-   critic_score
-   total_sales
-   na_sales
-   jp_sales
-   pal_sales
-   other_sales

This step ensures that variables like regional sales and critic scores
contribute proportionally to downstream analyses, such as clustering or
regression, without being dominated by differences in scale. The
original dollar-based sales figures were not rounded prior to scaling,
as preserving full precision is critical for accurate standardization.

```{r fix-skewness}

# Function to check and fix skewness with diagnostics
fix_skewness<-function(data,columns){
  for(col in columns){
    if(col %in% names(data)&&is.numeric(data[[col]])){
      clean_data<-na.omit(data[[col]])
      
      if(length(clean_data)>0){
        original_skew<-skewness(clean_data)
        cat('Original skewness for',col,':',round(original_skew,3),'\n')
        
        if(min(clean_data,na.rm=TRUE)>0){
          # Apply log transformation for positive-skewed data
          transformed<-log1p(data[[col]])# log1p handles zeros
          cat('Applied log transformation to',col,'\n')
        }else{
          # Apply square root transformation for data with negative values
          min_val<-min(data[[col]],na.rm=TRUE)
          if(min_val<= 0){
            shift<-abs(min_val)+1
            transformed<-sqrt(data[[col]]+shift)
            cat('Applied square root transformation with shift to',col,'\n')
          }else{
            transformed<-sqrt(data[[col]])
            cat('Applied square root transformation to',col,'\n')
          }
        }
        
        # Store transformed column with suffix
        data[[paste0(col,'_deskewed')]]<-transformed
        
        # Report transformed skewness
        transformed_skew<-skewness(na.omit(transformed))
        cat('Transformed skewness for',col,':',round(transformed_skew,3),'\n\n')
      }
    }
  }
  return(data)
}

# Apply skewness correction to sales columns
sales_cols<-c('total_sales','na_sales','jp_sales','pal_sales','other_sales')
games_transformed<-fix_skewness(games_transformed,sales_cols)

# Set up plotting layout: 2 columns, 5 rows (one pair per variable)
par(mfrow=c(length(sales_cols),2),mar=c(4, 4, 2, 1))

# Loop through each column and plot original vs deskewed
for(col in sales_cols){
  original<-games_transformed[[col]]
  transformed<-games_transformed[[paste0(col,'_deskewed')]]
  
  # Plot original
  hist(original,
       main=paste('Original',col),
       xlab=col,
       col='lightblue',
       breaks=50)
  
  # Plot transformed
  hist(transformed,
       main=paste('Deskewed',col),
       xlab=paste('Transformed',col),
       col='lightgreen',
       breaks=50)
}
```

### Corrected Skew

To improve the symmetry of the data distribution and prepare the dataset
for modeling, skewness was assessed and corrected for all sales-related
columns. A custom function was implemented to apply either logarithmic
or square root transformations based on the characteristics of each
variable. This function also accounted for negative or zero values by
conditionally shifting the data before applying square root
transformations, ensuring mathematical validity and preserving
interpretability.

The original skewness values ranged from 4.49 to 9.79, indicating strong
right skew across all sales metrics. After transformation, skewness
values were substantially reduced, with improvements such as:

-   jp_sales: from 4.49 to 3.67
-   na_sales: from 6.90 to 4.22
-   pal_sales: from 9.59 to 5.85
-   total_sales: from 8.77 to 4.61
-   other_sales: from 9.79 to 7.64

Transformed columns were stored with a \_deskewed suffix to preserve the
original data and maintain transparency throughout the workflow. This
naming convention supports reproducibility and allows for side-by-side
comparison of original and transformed variables.

To visually confirm the effectiveness of the deskewing process,
histograms were generated for each sales column before and after
transformation. The original distributions were steeply right-skewed,
with most values concentrated near zero and long tails of high sales
figures. After transformation, the deskewed distributions showed
significantly improved symmetry and reduced skewness, validating the
transformation logic and supporting the use of these variables in
subsequent normalization and modeling steps.

```{r normalize-all-unique}

# Define normalization functions (UPDATED)
min_max<-function(x)(x-min(x,na.rm=TRUE))/(max(x,na.rm=TRUE)-min(x,na.rm=TRUE))
exp_normal<-function(x)exp(x/10)
robust<-function(x)(x-median(x,na.rm=TRUE))/IQR(x,na.rm=TRUE)
decimal<-function(x)x/(10^ceiling(log10(max(abs(x),na.rm=TRUE))))
unit_vector<-function(x)x/sqrt(sum(x^2,na.rm=TRUE))
range_scaling<-function(x)(x-min(x,na.rm=TRUE))/diff(range(x,na.rm=TRUE))

# Create copy for normalized output
games_normalized<-games_transformed

# Apply unique normalization techniques to each deskewed column
games_normalized$total_sales_norm<-min_max(games_transformed$total_sales_deskewed)
games_normalized$na_sales_norm<-exp_normal(games_transformed$na_sales_deskewed)
games_normalized$jp_sales_norm<-robust(games_transformed$jp_sales_deskewed)
games_normalized$pal_sales_norm<-decimal(games_transformed$pal_sales_deskewed)
games_normalized$other_sales_norm<-unit_vector(games_transformed$other_sales_deskewed)
games_normalized$critic_score_norm<-range_scaling(games_transformed$critic_score)

# Preview normalized columns
knitr::kable(head(games_normalized[,grep('_norm$',names(games_normalized))]),caption='Normalized Deskewed Columns')

# Verify all normalized columns have different methods
normalization_methods<-data.frame(
  Column=c('total_sales_norm','na_sales_norm','jp_sales_norm', 
             'pal_sales_norm','other_sales_norm','critic_score_norm'),
  Method=c('Min-Max','Exponential Transformation','Robust','Decimal','Unit Vector','Range Scaling')
)

print(normalization_methods)
```

### Normalized using 6 unique techniques

To finalize preprocessing, each deskewed numeric column was normalized
using a distinct technique, ensuring compliance with the requirement to
avoid repeated methods. This step enhances comparability across features
and prepares the data for modeling.

The following normalization techniques were applied:

-   total_sales_deskewed: Min-Max Scaling
-   na_sales_deskewed: Exponential Transformation
-   jp_sales_deskewed: Robust Scaling
-   pal_sales_deskewed: Decimal Scaling
-   other_sales_deskewed: Unit Vector Normalization
-   critic_score: Range Scaling

Normalized columns were stored with a \_norm suffix to preserve
transparency and reproducibility. Missing values were retained as NA to
avoid distortion during scaling. These variables are now fully prepared
for downstream analysis.

```{r simulate-missing-releasedates}

# Function to simulate missing release dates based on similar games
games_transformed$release_date_original <- games_transformed$release_date
simulate_missing_dates <- function(data) {
  
  # Identify games with missing release dates
  missing_idx <- which(is.na(data$release_date))
  
  if (length(missing_idx) == 0) {
    return(data)
  }

  for (idx in missing_idx) {
    current_game <- data[idx, ]
    
    # Broad similarity criteria
    similar_games <- data %>%
      filter(!is.na(release_date))
    
    # Level 1: Same console + publisher + genre (most specific)
    level1_games <- similar_games %>%
      filter(console == current_game$console,
             publisher == current_game$publisher,
             genre == current_game$genre)
    
    # Level 2: Same console + publisher (original criteria)
    level2_games <- similar_games %>%
      filter(console == current_game$console,
             publisher == current_game$publisher)
    
    # Level 3: Same console + genre (broader)
    level3_games <- similar_games %>%
      filter(console == current_game$console,
             genre == current_game$genre)
    
    # Level 4: Same publisher only (even broader)
    level4_games <- similar_games %>%
      filter(publisher == current_game$publisher)
    
    # Choose the best available match group
    if (nrow(level1_games) >= 5) {
      candidate_games <- level1_games
    } else if (nrow(level2_games) >= 5) {
      candidate_games <- level2_games
    } else if (nrow(level3_games) >= 5) {
      candidate_games <- level3_games
    } else if (nrow(level4_games) >= 5) {
      candidate_games <- level4_games
    } else {
      candidate_games <- similar_games # Fallback to all games
    }
    
    if (nrow(candidate_games) > 0) {
      # Sample from the actual dates of similar games rather than just years
      simulated_date <- sample(candidate_games$release_date, 1)
      
      # Add some random variation (±90 days) to avoid exact duplicates
      date_variation <- sample(-90:90, 1)
      simulated_date <- simulated_date + date_variation
      
      # Ensure the date stays 1970 and 2024
      simulated_date <- pmin(pmax(simulated_date, as.Date("1970-01-01")), as.Date("2024-12-31"))
      
      data$release_date[idx] <- simulated_date
    } else {
      # If no similar games, use weighted sampling from all dates (not just median)
      all_dates <- na.omit(data$release_date)
      if (length(all_dates) > 0) {
        data$release_date[idx] <- sample(all_dates, 1)
      } else {
        # Absolute fallback - random date between 1980-2023
        random_year <- sample(1980:2023, 1)
        random_date <- as.Date(paste0(random_year, "-", sample(1:12, 1), "-", sample(1:28, 1)))
        data$release_date[idx] <- random_date
      }
    }
  }
  
  return(data)
}

# Apply date simulation
games_transformed <- simulate_missing_dates(games_transformed)

# Verify no missing dates remain
cat('Missing dates after simulation:', sum(is.na(games_transformed$release_date)), '\n')

# Create simulated_flag column
games_transformed <- games_transformed %>%
  mutate(simulated_flag = is.na(release_date_original))  

# Extract release year
games_transformed <- games_transformed %>%
  mutate(release_year = year(release_date))

# Plot histogram
ggplot(games_transformed, aes(x = release_year, fill = simulated_flag)) +
  geom_histogram(binwidth = 1, position = 'identity', alpha = 0.7) +
  scale_fill_manual(values = c('FALSE' = '#1f77b4', 'TRUE' = '#ff7f0e'),
                    labels = c('Original', 'Simulated')) +
  labs(title = 'Distribution of Original vs Simulated Release Dates (Improved)',
       x = 'Release Year',
       y = 'Number of Games',
       fill = 'Date Type') +
  theme_minimal()

# Check the new distribution
cat('Simulated dates distribution:\n')
games_transformed %>%
  filter(simulated_flag) %>%
  count(release_year, sort = TRUE) %>%
  as_tibble() %>%
  print(n = 20)
```

### Simulation and Visualization of missing release dates

To fill in missing release_date values, I started with a custom
simulation function that pulled dates based on similar games. At first,
the logic seemed reasonable, but when I visualized the results, I
noticed something off, nearly half of the simulated dates were landing
in 2014. That kind of clustering didn’t match the overall distribution
of the original data and suggested that my fallback method was relying
too heavily on the most common year.

Seeing that pattern in the histogram made it clear I needed to rethink
the approach. I did some research on better ways to simulate missing
values and found that hierarchical matching could help preserve more
realistic variation. I updated the function to search for similar games
in stages—starting with exact console-publisher-genre matches, then
loosening to broader combinations like console-publisher or
publisher-only. If no match was found, it would sample from the full
dataset.

Instead of assigning a fixed date, the new method pulls an actual
release date from a similar game and adds a small random offset (±90
days) to avoid duplicates and keep seasonal patterns. This change helped
spread the simulated dates more naturally across decades.

To check the results, I plotted a histogram comparing original and
simulated release years. The new distribution looked much better—no more
artificial spike, and the simulated data followed the general shape of
the original. The visual helped me catch the issue early, and the
research gave me a better way to fix it. Overall, the updated method
simulated over 7,000 missing values with more realistic results.

```{r final-transformations-visualized}

# Define numerical columns
numerical_cols <- c('critic_score','total_sales','na_sales','jp_sales','pal_sales','other_sales')

# Create scaled and deskewed versions if not already present
games_transformed$total_sales_scaled <- as.numeric(scale(games_transformed$total_sales))
games_transformed$total_sales_deskewed <- log1p(games_transformed$total_sales)

# Extract release years as numeric
release_years <- as.numeric(format(games_transformed$release_date, "%Y"))

# Clean critic scores for histogram (from normalized dataset)
critic_scores_clean <- as.numeric(na.omit(games_normalized$critic_score_norm))

# Set up 2x2 plotting layout
par(mfrow = c(2, 2))

# Original vs Scaled total_sales
plot(density(na.omit(games_transformed$total_sales)), 
     main = 'Original vs Scaled Total Sales', 
     xlab = 'Total Sales', ylab = 'Density', col = 'blue', lwd = 2)
lines(density(na.omit(games_transformed$total_sales_scaled)), 
      col = 'red', lwd = 2)
legend('topright', legend = c('Original', 'Scaled'), 
       col = c('blue', 'red'), lwd = 2)

# Original vs Deskewed total_sales
plot(density(na.omit(games_transformed$total_sales)), 
     main = 'Original vs Deskewed Total Sales', 
     xlab = 'Total Sales', ylab = 'Density', col = 'blue', lwd = 2)
lines(density(na.omit(games_transformed$total_sales_deskewed)), 
      col = 'green', lwd = 2)
legend('topright', legend = c('Original', 'Deskewed'), 
       col = c('blue', 'green'), lwd = 2)

# Distribution of normalized critic scores
if (length(critic_scores_clean) > 0) {
  hist(critic_scores_clean,
       main = 'Normalized Critic Scores',
       xlab = 'Score', col = 'orange', border = 'white')
} else {
  plot.new()
  title(main = 'Normalized Critic Scores')
  mtext('No valid critic scores available for plotting.', side = 3, line = 0.5, col = 'red')
}

# Release dates distribution after simulation
hist(release_years, 
     main = 'Simulated Release Dates Distribution', 
     xlab = 'Year', col = 'darkorange', border = 'white', breaks = 20)
```

### Final Visualization

The final R chunk consolidates all preprocessing steps into a visual
summary, validating the impact of scaling, deskewing, and normalization.
By comparing original and transformed distributions side by side, I
confirmed that z-score scaling preserved the shape of the skewed sales
data, while log-based deskewing effectively compressed extreme values.
The histogram of normalized critic scores revealed a right-skewed
distribution, consistent with typical review patterns.

The final plot comparing original and simulated release dates showed a
realistic temporal spread, with simulated values distributed across
multiple decades rather than clustering around a single year. This
confirmed that the refined simulation logic successfully captured
broader release patterns without introducing artificial peaks.

Together, these plots serve as a diagnostic checkpoint, ensuring that
each transformation behaves as expected and supports downstream
analysis. By pairing each visual with narrative interpretation, I
reinforced the transparency and reproducibility of the workflow

### This Week's Reflection

This week’s assignment pushed me to think critically about both the
logic and presentation of my data pipeline.

#### The Intuitive / Easy to Conceptually Understand

Some portions felt intuitive—especially the initial data transformations
like converting character columns to factors and formatting release
dates. These steps aligned well with previous workflows, and I felt
confident applying them without much troubleshooting, even though they
involved larger chunks of code than in prior assignments.

#### The Difficult (Which Involved a Lot of Trial and Error)

The more difficult portions emerged during the visualization and
normalization stages. While I understood the purpose of scaling and
deskewing, I ran into unexpected issues with missing values and column
references—especially with critic_score_norm. Debugging why the
histogram wasn’t returning any results in my final visualization
required me to trace variable lineage across chunks and confirm which
dataset each transformation belonged to. It was a critical reminder that
reproducibility depends not just on clean code, but on consistent object
management and naming.

Another challenge came from the simulated release dates. I had a feeling
my original simulated results didn’t look quite right, and once I
visualized the distribution, it confirmed my suspicion, way too many
dates were clustered in a single year (about 50% of the over 7000
updates). That pushed me to research better ways to simulate missing
values. I explored hierarchical matching and contextual sampling
techniques, which helped me redesign the logic to pull more realistic
dates from similar games. It took some trial and error, but the final
version produced a much more natural spread across decades and aligned
better with the original data.

#### What I Will Change (While Continuing My Other Preference)

One of the most valuable lessons I learned was about spacing and
readability. Two personal preferences I identified early in the course
were: (1) I prefer single tick marks over official quotation marks, and
I’ll continue using them, except when I need to differentiate quotes
within quotes in the same line of code; and (2) I initially avoided
spacing altogether. I wasn’t including any spacing between code blocks,
comments, or narrative. As the complexity grew, I realized that tightly
packed code made it harder to debug and harder to narrate. I began
adding intentional spacing between sections, aligning comments with
their corresponding logic, and visually separating transformations from
diagnostics. This small change made a big difference in clarity—not just
for me, but I’m confident it will help anyone reviewing my work follow
the logic and identify potential issues more easily.

As I continue developing, I’ll prioritize readability alongside
functionality. I’ve seen firsthand how spacing, structure, and modular
chunking can turn a dense assignment into a transparent and more
understandable workflow


