---
title: "HollisT_Week7"
author: "Tim Hollis"
date: "`r Sys.Date()`"
output: html_document
---

```{r descriptive-stats, results='asis'}

#data

with_tv <- c(12, 7, 3, 11, 8, 5, 14, 7, 9, 10)
without_tv <- c(8, 7, 4, 14, 6, 7, 12, 5, 5, 8)

# Mean

mean_with_tv <- mean(with_tv)
mean_without_tv <- mean(without_tv)

# SD

sd_with_tv <- sd(with_tv)
sd_without_tv <- sd(without_tv)

# Variance

var_with_tv <- var(with_tv)
var_without_tv <- var(without_tv)
```

### **R-Markdown, then R-Chunk summary tables**

| Condition     | Mean                | SD                  | Variance             |
|:-------------:|:------------------:|:-------------------:|:--------------------:|
| With TV       | `r mean_with_tv`    | `r sd_with_tv`       | `r var_with_tv`      |
| Without TV    | `r mean_without_tv` | `r sd_without_tv`    | `r var_without_tv`   |

```{r summary-table}
# Create the summary table

stats_table <- data.frame(
  Condition = c('With TV', 'Without TV'),
  Mean = c(mean_with_tv, mean_without_tv),
  SD = c(sd_with_tv, sd_without_tv),
  Variance = c(var_with_tv, var_without_tv)
)

# Render with borders and centered alignment

kable(stats_table, align = 'c') %>%
  kable_styling(full_width = FALSE, position = 'center', bootstrap_options = c('striped', 'bordered'))
```


#### **Summary**

The descriptive statistics suggest that children consumed minimally more candy 
corn on average after watching TV (`r mean_with_tv`) compared to without TV 
(`r mean_without_tv`). The variability in both conditions is relatively similar,
as shown by the standard deviations and variances.

To determine whether this observed difference is statistically significant, we 
will now conduct a paired-sample t-test.

### **Paired t-test Analysis**

```{r paired-t-test}

# Conduct paired t-test

t_test_result <- t.test(with_tv, without_tv, paired = TRUE)
t_test_result
```

#### **Summary: Interpretation of t-test Results**

- **Null Hypothesis (H₀):** There is no difference in candy corn consumption 
between the TV and no-TV conditions.
- **Alternative Hypothesis (H₁):** Candy corn consumption increases after TV 
stimulation.
- **Significance Level (α):** 0.05, chosen as a conventional threshold for 
detecting meaningful effects.
- **Degrees of Freedom:** `r t_test_result$parameter`
- **t-statistic:** `r round(t_test_result$statistic, 3)`
- **p-value:** `r round(t_test_result$p.value, 4)`

Since the p-value is `r round(t_test_result$p.value, 4)`, which is greater than 
0.05, we **fail to reject the null hypothesis**. This suggests that TV 
stimulation **does not have a statistically significant effect** on candy corn 
consumption. While there's a small observed difference in means (8.6 vs 7.6), 
this difference is not statistically significant and could easily occur by 
random chance

### **Visual Comparison**


```{r boxplot-visual}

# data
means <- c(mean(with_tv), mean(without_tv))
overall_mean <- mean(c(with_tv, without_tv))

# Create the boxplot
boxplot(with_tv, without_tv,
        names = c('With TV', 'Without TV'),
        col = c('orange', 'lightblue'),
        main = 'Candy Corn Consumption by Condition',
        ylab = 'Pieces Eaten',
        ylim = c(0, 20))


# Add dashed line for overall mean
abline(h = overall_mean, col = 'darkgreen', lty = 2)

# Add mini legend in top right
legend('topright',
       legend = c('Group Mean', 'Overall Mean'),
       col = c('black', 'darkgreen'),
       lty = c(1, 2),
       pch = c(NA, NA),
       bty = 'n',
       cex = 0.8)
```


#### **Summary**

The paired t-test revealed no significant increase in candy corn consumption 
after TV exposure. This fails to support the researchers assumption that TV 
exposure has a significant impact on candy corn consumption.



## **Reflection on Week’s Learning**

**What felt intuitive:** 

- Setting up the data vectors and calculating descriptive statistics felt 
straightforward and reinforced my understanding of basic R functions.

**What was challenging:** 

- Interpreting the t-test output and remembering the assumptions of paired tests
required a bit more review. I will be honest, I originally interpreted the 
alternate hypothesis as being correct based on the difference between the means 
being 1 and not 0, but had to reword and resubmit my assignment when I finally
factored in the p-value of .22 being higher than .05, so theres a 22% chance, 
the means being greater than 0, is random and based on sample size!

**What I’ll stop doing:** 

- Relying solely on visual inspection without statistical testing, the t-test 
added clarity and confidence to the interpretation. I will also ensure that my
interpretation of data is inclusive of all factors included and not jump to a 
conclusion based on one factor.

**What I’ll continue doing:** 

- Using both Markdown and styled R tables for clarity, and incorporating visuals
to support statistical findings.

