---
title: "HollisT_Week8"
author: "Tim Hollis"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)

# Load required libraries

library(dplyr)
library(ggplot2)
library(corrplot)
library(lsa)
library(viridis)
library(purrr)


# Remove scientific notation globally

options(scipen = 999)  

# Set global theme for all ggplot visualizations

theme_set(theme_minimal())

# Set global theme modifications

theme_update(
  plot.title = element_text(hjust = 0.5, face = "bold"),
  axis.title = element_text(face = "bold"),
  legend.position = "none"
)

# Load Dataset

travel_data <- read.csv('travel_insurance.csv')
```
```{r data-view-clean}

# Examine the initial data structure

glimpse(travel_data)
head(travel_data)

cat("Rows with negative net sales:", sum(travel_data$Net.Sales < 0, na.rm = TRUE), "\n")
cat("Rows with zero net sales:", sum(travel_data$Net.Sales == 0, na.rm = TRUE), "\n")
cat("Rows removed due to Age > 100:", sum(travel_data$Age > 100, na.rm = TRUE), "\n")

# Overlap check

cat("Rows removed for both conditions:", sum(travel_data$Net.Sales <= 0 & travel_data$Age > 100, na.rm = TRUE), "\n")

# Total unique rows to be removed

cat("Total rows to be removed:", sum(travel_data$Net.Sales <= 0 | travel_data$Age > 100, na.rm = TRUE), "\n")
cat("Original total rows:", nrow(travel_data), "\n")


# Clean the data: keep only rows with Net Sales > 0 AND fix the spelling

clean_data <- travel_data %>% 
  filter(Net.Sales > 0, Age <= 100) %>% # Filtering out rows with net sales data that is not positive, and age over 100
  rename(Commission = Commision..in.value.) # Correcting spelling of Commission


cat("Rows after cleaning:", nrow(clean_data), "\n")
cat("Percentage of data retained:", round(nrow(clean_data)/nrow(travel_data)*100, 2), "%\n")
cat("Column names after cleaning:", names(clean_data), "\n")
```
<hr style="border: 1px solid #000000;"> 
  
#### **Data Integrity/Cleaning Results**

Out of **63,326** total rows:

- **678** entries contain negative Net Sales values, likely due to refunds, 
cancellations, or data entry errors.
- **1,884** entries have zero Net Sales, possibly incomplete or placeholder 
transactions.
- **984** entries list an age over 100, which may reflect data entry issues or 
unrealistic values.
- **31** entries meet more than one condition, so we subtract this overlap to 
avoid double-counting.

After removing **3,515** rows, **59,811** valid entries remain, approximately 
**94.45%** of the original dataset. These cleaned records are suitable for analysis.

<hr style="border: 3px solid #000000;">

## **Assignment Overview**

This week’s assignment explores two core statistical concepts: correlation and 
confidence intervals. Using the travel insurance dataset, we’ll compute various 
correlation metrics between variables, visualize relationships, and interpret 
the strength and direction of associations. We’ll then shift to estimating 
confidence intervals and t-scores for numeric features to understand variability
and inference.

<hr style="border: 1px solid #000000;">

### **Correlation Coefficient Analysis**

We begin by exploring the relationship between **Age** and **Net Sales** as the 
first correlation analysis, and my original assumption is there is not going
to be a strong correlation, I chose two variables unrelated to each other in general, 
so will be surprised if I am incorrect.

```{r basic-correlation}


ggplot(clean_data, aes(x = Age, y = Net.Sales, color = Age)) +
  geom_point(alpha = 0.4) +
  scale_color_viridis_c(option = "C") +
  labs(
    title = "Scatterplot of Age and Net Sales",
    x = "Age",
    y = "Net Sales",
    color = "Age"
  ) +
  theme(plot.title = element_text(hjust = 0.5, face = "bold"))


# Compute Pearson correlation between Age and Net Sales

cor_age_net_sales <- cor(clean_data$Age, clean_data$Net.Sales, method = "pearson", use = "complete.obs")

cor_age_net_sales
```

<hr style="border: 1px solid #000000;">

#### **Summary**

The computed correlation coefficient between Age and Net Sales is `r round(cor_age_net_sales, 4)`, 
indicating a `r ifelse(abs(cor_age_net_sales) < 0.3, "weak", ifelse(abs(cor_age_net_sales) < 0.7, "moderate", "strong"))` 
`r ifelse(cor_age_net_sales > 0, "positive", "negative")` relationship.

This aligns with the initial hypothesis: Age and Net Sales are not strongly 
associated in this dataset. The weak correlation suggests that age is not a 
meaningful predictor of sales performance, reinforcing the assumption that 
these variables operate independently in this context.

<hr style="border: 1px solid #000000;">

### **Pearson, Spearman, and Kendall Comparison**

 While age provides interesting demographic insights, the relationship between 
 **commission** structures and **sales** performance is often more directly relevant for
 business strategy. Let's now examine how **commission** values correlate with 
 **net sales** using multiple correlation methods to see how the Pearson, Spearman, and
 Kendall Correlations compare when analyzing the relationship betweem those variables, that
 I assume will be strongly positively correlated.
 
```{r P-S-K}

# Compute different correlation methods between Commission and Net Sales

pearson_cor <- cor(clean_data$Commission, clean_data$Net.Sales, method = 'pearson', use = 'complete.obs')
spearman_cor <- cor(clean_data$Commission, clean_data$Net.Sales, method = 'spearman', use = 'complete.obs')
kendall_cor <- cor(clean_data$Commission, clean_data$Net.Sales, method = 'kendall', use = 'complete.obs')

# Create comprehensive correlation methods table

correlation_methods_table <- data.frame(
  Correlation_Type = c('Pearson', 'Spearman', 'Kendall'),
  Best_For_Measuring = c(
    'Linear relationships between continuous variables',
    'Monotonic relationships (linear or nonlinear)',
    'Concordance and rank-based associations'
  ),  
  Correlation_Value = c(round(pearson_cor, 4), round(spearman_cor, 4), round(kendall_cor, 4)),
  Strength_Interpretation = c(
    ifelse(abs(pearson_cor) < 0.3, 'Weak', ifelse(abs(pearson_cor) < 0.7, 'Moderate', 'Strong')),
    ifelse(abs(spearman_cor) < 0.3, 'Weak', ifelse(abs(spearman_cor) < 0.7, 'Moderate', 'Strong')),
    ifelse(abs(kendall_cor) < 0.3, 'Weak', ifelse(abs(kendall_cor) < 0.7, 'Moderate', 'Strong'))
  )
)

# Display the table

knitr::kable(correlation_methods_table, caption = 'Commission and Net Sales: Comparison of Correlation Methods')
```

<hr style="border: 1px solid #000000;">

#### **Summary**

All three correlation methods reveal a `r ifelse(mean(c(pearson_cor, spearman_cor, kendall_cor)) > 0.7, "strong", "moderate")` positive relationship between Commission and Net Sales. The Pearson correlation (`r round(pearson_cor, 4)`) suggests the strongest linear association, while Spearman and Kendall confirm that the relationship holds across ranked and monotonic patterns.

This supports the hypothesis that higher commissions are closely tied to higher 
net sales, a finding that aligns with business logic and reinforces the importance of 
incentive structures in sales performance.

<hr style="border: 1px solid #000000;">

### **Correlation Matrix for All Variables**

Now that we've established strong commission-sales relationships, let's expand our view to examine
how all numeric variables interact by creating a comprehensive correlation matrix.

```{r correlation-matrix-all}

# Select only numeric variables for correlation matrix

numeric_data <- clean_data %>% 
  select(where(is.numeric)) %>% 
  select(-matches('NA', ignore.case = TRUE))

# Compute correlation matrix

correlation_matrix <- cor(numeric_data, use = 'complete.obs')

# Create visualization with values in boxes

corrplot(correlation_matrix, method = 'color', type = 'upper', 
         order = 'hclust', tl.cex = 0.8, tl.col = 'black',
         title = 'Correlation Matrix of Numeric Variables',
         mar = c(0, 0, 1, 0),
         addCoef.col = 'black',
         number.cex = 0.7,
         col = colorRampPalette(c('#2166ac', '#f7f7f7', '#b2182b'))(100))

# Display top correlations excluding self-correlations and duplicates

correlation_pairs <- which(correlation_matrix > 0.3 & correlation_matrix < 1, arr.ind = TRUE)
key_relationships <- data.frame(
  Variable_1 = rownames(correlation_matrix)[correlation_pairs[, 1]],
  Variable_2 = colnames(correlation_matrix)[correlation_pairs[, 2]],
  Correlation = round(correlation_matrix[correlation_pairs], 3)
) %>% 
  filter(Variable_1 < Variable_2) %>% 
  arrange(desc(Correlation))

# Get strongest unique relationships

top_relationships <- head(key_relationships, 3)

knitr::kable(top_relationships, caption = 'Strongest Variable Relationships (Correlation > 0.3)')
```

<hr style="border: 1px solid #000000;">

#### **Summary**

The correlation matrix reveals `r nrow(top_relationships)` strong relationships 
between numeric variable pairs. The strongest correlation is between `r top_relationships$Variable_1[1]` 
and `r top_relationships$Variable_2[1]`, with a value of `r top_relationships$Correlation[1]`.

This confirms the earlier hypothesis: **Net Sales** and **Commission** are tightly linked, 
forming the most robust relationship in the dataset. The matrix also highlights 
other meaningful associations worth exploring in future analyses, such as duration or age-based patterns.

<hr style="border: 1px solid #000000;">

### **Covariance: Commission and Net Sales**

Beyond correlation, understanding how variables co-vary provides additional insights 
into their relationships. Let's compute the covariance between Commission and Net Sales.

```{r cov-commission-netsales}

# Compute covariance between Commission and Net Sales

cov_commission_net_sales <- cov(clean_data$Commission, clean_data$Net.Sales, use = 'complete.obs')

# Create covariance summary

covariance_summary <- data.frame(
  Variable_1 = 'Commission',
  Variable_2 = 'Net Sales',
  Covariance = round(cov_commission_net_sales, 2),
  Interpretation = ifelse(cov_commission_net_sales > 0, 
                        'Positive relationship: variables move together',
                        'Negative relationship: variables move oppositely'),
  Magnitude = ifelse(abs(cov_commission_net_sales) > 1000, 'Large', 
                    ifelse(abs(cov_commission_net_sales) > 100, 'Moderate', 'Small'))
)

knitr::kable(covariance_summary, caption = 'Covariance Analysis: Commission and Net Sales')
```

<hr style="border: 1px solid #000000;">

#### **Summary**

The covariance of `r round(cov_commission_net_sales, 2)` indicates a `r ifelse(cov_commission_net_sales > 0, 'positive', 'negative')` relationship with `r ifelse(abs(cov_commission_net_sales) > 1000, 'large', ifelse(abs(cov_commission_net_sales) > 100, 'moderate', 'small'))` magnitude, confirming that Commission and Net Sales tend to move together.

<hr style="border: 1px solid #000000;">

### **Cosine Similarity between Commission and Net Sales**

Finally, let's compute cosine similarity to measure the angular relationship.

```{r Cosine-Commission-NetSales}

# Clean and select relevant columns

comm_net_clean <- clean_data %>%
  select(Commission, Net.Sales) %>%
  na.omit()

# Compute correlation metrics

pearson_cor <- cor(comm_net_clean$Commission, comm_net_clean$Net.Sales, method = 'pearson')
spearman_cor <- cor(comm_net_clean$Commission, comm_net_clean$Net.Sales, method = 'spearman')
kendall_cor <- cor(comm_net_clean$Commission, comm_net_clean$Net.Sales, method = 'kendall')

# Manual cosine similarity calculation 

x <- comm_net_clean$Commission
y <- comm_net_clean$Net.Sales

# Remove any infinite or NA values

x <- x[is.finite(x) & is.finite(y)]
y <- y[is.finite(x) & is.finite(y)]

# Calculate cosine similarity manually

cosine_value <- sum(x * y) / (sqrt(sum(x^2)) * sqrt(sum(y^2)))

# Display result

round(cosine_value, 4)
```

<hr style="border: 1px solid #000000;">

#### **Summary**

The cosine similarity between **Commission** and **Net Sales** is approximately
`r round(cosine_value, 4)`, indicating a strong directional alignment. This 
suggests that when net sales increase, commission values tend to increase in a 
consistent pattern, even if their absolute magnitudes differ.

Unlike correlation, cosine similarity focuses on the angle between vectors 
rather than their scale, making it especially useful when comparing 
proportional trends across variables.

<hr style="border: 1px solid #000000;">

#### **Scatterplot and Similarity Comparison**

```{r similarity-plot}

# Create a similarity comparison data frame

similarity_scores <- data.frame(
  Method = c('Pearson', 'Spearman', 'Kendall', 'Cosine'),
  Score = c(pearson_cor, spearman_cor, kendall_cor, cosine_value)  
)

# Show Columns in decending order

similarity_scores$Method <- factor(similarity_scores$Method, levels = similarity_scores$Method[order(similarity_scores$Score, decreasing = TRUE)])

# Plot similarity scores

ggplot(similarity_scores, aes(x = Method, y = Score, fill = Method)) +
  geom_col(width = 0.6, show.legend = FALSE) +
  geom_text(aes(label = round(Score, 3)), vjust = -0.5, size = 4.5) +
  ylim(0, 1) +
  labs(
    title = 'Similarity Metrics: Commission and Net Sales',
    x = 'Method',
    y = 'Similarity Score'
  ) +
  theme(
    plot.title = element_text(hjust = 0.5, face = 'bold'),
    axis.title = element_text(face = 'bold')
  )
```

<hr style="border: 1px solid #000000;">

#### **Summary: Comparing Similarity Metrics**

This analysis compares four similarity measures between **Commission** and **Net Sales**:

- **Pearson** captures linear correlation.
- **Spearman** and **Kendall** assess rank-based monotonic relationships.
- **Cosine similarity** evaluates directional alignment between vectors.

In this case:

- Cosine similarity (`r round(cosine_value, 3)`) was notably strong, suggesting
proportional alignment.

- Pearson (`r round(pearson_cor, 3)`) confirmed a linear relationship.

- Spearman (`r round(spearman_cor, 3)`) and Kendall (`r round(kendall_cor, 3)`) 
showed consistent rank-order association.

Cosine adds a complementary perspective by focusing on **vector orientation**, 
not just value distribution.

<hr style="border: 1px solid #000000;">

### **Comparison of All Discussed Measures of Similarity**

A combined view of the measurements of correlation, similarity, direction, etc.
for Sales and Commission variables.

```{r all-in-comparison}

# Create comparison table of all similarity measures

similarity_comparison <- data.frame(
  Method = c('Pearson', 'Spearman', 'Kendall', 'Cosine', 'Covariance'),
  Value = c(pearson_cor, spearman_cor, kendall_cor, cosine_value, cov_commission_net_sales), 
  Measurement_Type = c('Correlation', 'Correlation', 'Correlation', 'Similarity', 'Covariance'),
  Best_For = c(
    'Linear relationships',
    'Monotonic relationships', 
    'Rank concordance',
    'Angular similarity',
    'Joint variability'
  )
) %>% mutate(Value = round(Value, 4))

knitr::kable(similarity_comparison, caption = 'Comparison of All Similarity Measures: Commission and Net Sales')
```

<hr style="border: 1px solid #000000;">

#### **Summary**

Across all measures, including **Pearson**, **Spearman**, **Kendall**, 
**cosine similarity**, and **covariance**, we observe consistently strong 
positive associations between **Net Sales** and **Commission**.

This convergence across statistical and geometric metrics reinforces the 
reliability of their relationship. Whether viewed through linear correlation, 
rank concordance, angular similarity, or joint variability, the connection 
between these two variables remains robust — a key insight for incentive modeling
and sales strategy.

<hr style="border: 1px solid #000000;">

### **Confidence Intervals Analysis**

Having completed our correlation analysis, let's now shift to confidence intervals 
to understand the precision and reliability of our numeric feature estimates, 
beginning with the average Confidence Interval (CI) for each numeric variable.

```{r avg-ci}

# Define the compute_ci function 

compute_ci <- function(x, confidence = 0.95) {
  if (length(na.omit(x)) < 2) return(c(NA, NA, NA))
  
  n <- length(na.omit(x))
  mean_val <- mean(x, na.rm = TRUE)
  se <- sd(x, na.rm = TRUE) / sqrt(n)
  t_value <- qt((1 + confidence) / 2, df = n - 1)
  margin_error <- t_value * se
  
  return(c(
    lower = mean_val - margin_error,
    mean = mean_val,
    upper = mean_val + margin_error
  ))
}

# Compute CIs for all numeric variables using map_df

ci_results <- map_df(names(numeric_data), function(col_name) {
  ci <- compute_ci(numeric_data[[col_name]])
  data.frame(
    Variable = col_name,
    Mean = ci['mean'],
    CI_Lower = ci['lower'],
    CI_Upper = ci['upper'],
    CI_Width = ci['upper'] - ci['lower']
  )
})

# Create comprehensive precision table 

precision_table <- ci_results %>%
  mutate(
    Mean = round(Mean, 2),
    CI_Lower = round(CI_Lower, 2),
    CI_Upper = round(CI_Upper, 2),
    CI_Width = round(CI_Width, 2),
    Relative_Precision = round((CI_Width / Mean) * 100, 1),
    Precision_Rating = case_when(
      Relative_Precision < 1 ~ 'Very High',
      Relative_Precision < 3 ~ 'High',
      Relative_Precision < 5 ~ 'Medium',
      TRUE ~ 'Low'
    )
  ) %>%
  select(Variable, Mean, CI_Lower, CI_Upper, CI_Width, Relative_Precision, Precision_Rating)

# Display precision table

knitr::kable(precision_table, caption = 'Precision Analysis: 95% Confidence Intervals for Numeric Variables')
```

<hr style="border: 1px solid #000000;">

#### **Summary**

Precision ratings are based on relative precision (CI width as a percentage of the mean):

- **Very High**: < 1%
- **High**: 1–3%
- **Medium**: 3–5%
- **Low**: > 5%

The 95% confidence intervals (CI's) for each numeric variable reveal the 
precision of their mean estimates. Narrower intervals and lower relative 
precision indicate greater reliability and less variability in the data.

- **Age** exhibits the highest precision, with a relative precision of **0.6%**,
suggesting a highly stable mean estimate.

- **Net Sales** also shows high precision at **1.8%**, reinforcing its reliability.

- **Duration** and **Commission** fall into the medium precision range (~**3.3%**),
indicating slightly more variability but still acceptable consistency.

Overall, two variables (Age and Net Sales) demonstrate high or very high 
precision, while the remaining two are close behind — reinforcing the dataset’s 
overall quality and suitability for inferential analysis.

<hr style="border: 1px solid #000000;">

### **T-Scores**

Now let's compute t-scores to determine the statistical significance of each numeric 
feature's mean value.

```{r t-scores}

# Define the compute_t_scores function (without p-values)

compute_t_scores <- function(x) {
  if (length(na.omit(x)) < 2) return(NA)
  
  mean_val <- mean(x, na.rm = TRUE)
  se <- sd(x, na.rm = TRUE) / sqrt(length(na.omit(x)))
  t_score <- mean_val / se
  
  return(t_score)
}

# Compute t-scores for all numeric variables

t_score_results <- data.frame(
  Variable = names(numeric_data),
  T_Score = sapply(numeric_data, compute_t_scores),
  Mean_Value = sapply(numeric_data, function(x) mean(x, na.rm = TRUE)),
  Standard_Error = sapply(numeric_data, function(x) sd(x, na.rm = TRUE) / sqrt(length(na.omit(x))))
) %>% mutate(
  T_Score = round(T_Score, 2),
  Mean_Value = round(Mean_Value, 2),
  Standard_Error = round(Standard_Error, 4),
  Reliability = case_when(
    abs(T_Score) > 100 ~ "Very High",
    abs(T_Score) > 50 ~ "High", 
    abs(T_Score) > 10 ~ "Medium",
    TRUE ~ "Low"
  )
)

# Display t-score results

knitr::kable(t_score_results, caption = 'T-scores and Estimate Reliability Analysis')
```

<hr style="border: 1px solid #000000;">

#### **Final Summary**

This analysis explored the relationships, variability, and statistical 
significance of key numeric features in the travel insurance dataset using a 
multi-method approach:

**Correlation & Similarity**

- **Net Sales** and **Commission** showed consistently strong positive 
relationships across all correlation methods (Pearson, Spearman, Kendall),
with a cosine similarity of **0.79**, reinforcing their directional alignment.

- **Age** and **Net Sales** showed a weak correlation, confirming the assumption 
that age is not a strong predictor of sales performance.

**Covariance & Cosine**

- The positive **covariance** between **Net Sales** and **Commission** confirmed 
their tendency to increase together.

- **Cosine similarity** added a geometric perspective, showing that the two 
variables move in the same direction even if their scales differ.

**Confidence Intervals & Precision**

- **Age** had the tightest confidence interval and lowest relative precision, 
indicating the most stable mean estimate.

- **Net Sales** also demonstrated high precision, while **Duration** and **Commission** 
were slightly more variable but still reliable.

- All variables had **95% confidence intervals** that support the robustness of
their mean estimates.

**T-Scores & Significance**

- All four variables had very high t-scores, indicating that their means are 
statistically significant and practically meaningful.

- **Age** again stood out with a t-score over **900**, reinforcing its stability
and precision.

<hr style="border: 2px solid #000000;">

## **Reflection for Week 8**

<hr style="border: 1px solid #000000;">

### **Intuitive and Successful Components**

- Correlation coefficients between commission and net sales proved highly intuitive,
with the strong positive relationship clearly demonstrating that higher commissions 
correlate with increased sales performance

- Multiple correlation methods (Pearson, Spearman, Kendall) provided complementary
perspectives that reinforced the same business insight from different statistical angles

- Data cleaning was straightforward once recognizing that negative and zero net 
sales values represented invalid business cases requiring removal

- Cosine similarity calculations, while mathematically complex, produced results 
that aligned with correlation findings, validating the overall analysis approach

<hr style="border: 1px solid #000000;">

#### **Technical Successes**:

- Pushed myself to do more inline R code in R Markdown to dynamically reference 
calculated values with the tick marks at beginning of any inline calculation like the
***r round(cor_age_net_sales, 3)*** using the calculated values to show correlation 
between age and sales etc. instead of manually typing out the value

- Pushed myself to integrate if/elif type statements that complimented my recent
Python learnings

- Used HTML coding in R Makdown to create thicker borders between concepts, than 
the 3 dashes create

<hr style="border: 1px solid #000000;">

### **Challenging and Less Intuitive Aspects**

- Confidence interval visualizations proved counterproductive when comparing 
variables with different scales, as the graph attempted to display incomparable
metrics on a single axis

- Statistical significance interpretation required careful consideration of what 
constitutes meaningful business impact versus mathematical significance

- Correlation matrix interpretation initially overwhelmed with information until
focusing on the strongest relationships through filtering and ranking

<hr style="border: 1px solid #000000;">

#### **Technical Difficulties**:

- Debugging inline R Markdown syntax errors, particularly with proper backtick 
placement around dynamic values

- Selecting appropriate color schemes for correlation matrices that maintained 
readability while conveying meaningful color coding

- R Studio continued to get stuck at roughly the 64% knit completion, until I
realized I was using an incorrect command to calculate cosine similarity and then
manually calculated the relevant values

<hr style="border: 1px solid #000000;">

### What I will Stop, and What I will Continue Doing:

<hr style="border: 1px solid #000000;">

**STOP**:

- Creating misleading visualizations that compare variables with fundamentally 
different scales and units

- Overcomplicating analyses with unnecessary visualizations when clear tables suffice

- Using inline R code just for the sake of using inline code, using it when it adds
value dynamically, but typing a direct value when it does not.

<hr style="border: 1px solid #000000;">

**Continue**:

- Using multiple correlation methods to validate relationship findings from 
different statistical perspectives

- Implementing comprehensive data cleaning pipelines that document record removal 
rationale

- Creating clear, business-focused interpretations of statistical results

- Leveraging R Markdown's dynamic reporting capabilities to create reproducible analyses

<hr style="border: 3px solid #000000;">

