---
title: "HollisT_Week9"
author: "Tim Hollis"
date: "`r Sys.Date()`"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE, warning = FALSE, message = FALSE)

#Loading Libraries

library(tidyverse)
library(broom)
library(ggplot2)
library(knitr)
library(kableExtra)
library(gridExtra)
library(moments)

# Setting Theme

theme_custom <- theme_minimal() +
  theme(plot.title = element_text(face = 'bold', size = 14),
        axis.title = element_text(face = 'bold'),
        legend.position = 'bottom')

# Remove scientific expression for numerical data

options(scipen = 999)

# Set Seed for Reproducibility

set.seed(123)
```
```{r data-review}

# Load and inspect data

travel_data <- read.csv('travel_insurance_anova.csv')

# Convert factor variables to maintain proper ANOVA structure

travel_data$Agency.Type <- as.factor(travel_data$Agency.Type)
travel_data$Distribution.Channel <- as.factor(travel_data$Distribution.Channel)

# Examine the structure

cat('Structure of Data:\n')
str(travel_data)

# Check for missing values

missing_summary <- sapply(travel_data, function(x) sum(is.na(x)))
cat('Missing values in each column:')
print(missing_summary)

# Check the distribution of categorical variables

cat('Distribution of Agency Types:')
agency_dist <- table(travel_data$Agency.Type)
print(agency_dist)

cat('Distribution of Distribution Channels:')
channel_dist <- table(travel_data$Distribution.Channel)
print(channel_dist)

# Check proportions for both factors

cat('Proportions of Agency Types:')
agency_prop <- prop.table(agency_dist)
agency_prop_formatted <- data.frame(
  Agency_Type = names(agency_prop),
  Count = as.numeric(agency_dist),
  Proportion = paste0(round(agency_prop * 100, 2), '%')
)

cat('Proportions of Distribution Channels:')
channel_prop <- prop.table(channel_dist)
channel_prop_formatted <- data.frame(
  Channel = names(channel_prop),
  Count = as.numeric(channel_dist),
  Proportion = paste0(round(channel_prop * 100, 2), '%')
)

# Display both tables 

kable(agency_prop_formatted, caption = 'Agency Type') %>%
  kable_styling(bootstrap_options = c('striped', 'hover', 'condensed'), full_width = FALSE)

kable(channel_prop_formatted, caption = 'Distribution Channel') %>%
  kable_styling(bootstrap_options = c('striped', 'hover', 'condensed'), full_width = FALSE)

# Create a cross-tabulation

cross_tab <- table(travel_data$Agency.Type, travel_data$Distribution.Channel)
cat('Cross-tabulation of Agency Type vs Distribution Channel:')
print(cross_tab)

# Summary statistics by groups (all four possible combinations)

group_summary <- travel_data %>%
  group_by(Agency.Type, Distribution.Channel) %>%
  summarise(
    Count = n(),
    Mean_Net_Sales = mean(Net.Sales),
    SD_Net_Sales = sd(Net.Sales),
    Min_Net_Sales = min(Net.Sales),
    Max_Net_Sales = max(Net.Sales),
    .groups = 'drop'
  )

cat('Summary Statistics by Group:')
group_summary %>%
  mutate(across(c(Mean_Net_Sales, SD_Net_Sales, Min_Net_Sales, Max_Net_Sales), ~ paste0('$', round(., 2)))) %>%
  kable(caption = 'Net Sales Summary by Agency Type and Distribution Channel') %>%
  kable_styling(bootstrap_options = c('striped', 'hover', 'condensed'), full_width = FALSE)
```
<hr style="border: 1px solid #000000;">

### **Data Review Summary**

The dataset contains `r nrow(travel_data)` observations with two categorical
factors (Agency Type and Distribution Channel) and one continuous outcome variable
(Net Sales). There are no missing values, which is excellent for our analysis.
Both factors have two categories:

**Agency Type**

- ***Airlines*** (`r agency_dist['Airlines']` observations, `r round(prop.table(agency_dist)['Airlines'] * 100, 2)`%)
- ***Travel Agency*** (`r agency_dist['Travel Agency']` observations, `r round(prop.table(agency_dist)['Travel Agency'] * 100, 2)`%)

**Distribution Channel**

- ***Online*** (`r channel_dist['Online']` observations, `r round(prop.table(channel_dist)['Online'] * 100, 2)`%)
- ***Offline*** (`r channel_dist['Offline']` observations, `r round(prop.table(channel_dist)['Offline'] * 100, 2)`%)

**Note**: There is significant imbalance in the Distribution Channel factor, with 
**Online** channels representing `r round(prop.table(channel_dist)['Online'] * 100, 2)`% 
of observations. The Agency Type factor shows `r ifelse(prop.table(agency_dist)['Airlines'] > 0.6, "a moderate imbalance", "relatively balanced distribution")` 
(`r round(prop.table(agency_dist)['Airlines'] * 100, 2)`% Airlines vs `r round(prop.table(agency_dist)['Travel Agency'] * 100, 2)`% Travel Agencies). 
This imbalance should be considered when interpreting the effects of these factors.
<hr style="border: 1px solid #000000;">

### **Data Visualization**

The boxplots visualize the distribution of Net Sales across the categories of 
the independent variables.

```{r data-viz}

# Boxplot for Agency Type

p1 <- ggplot(travel_data, aes(x = Agency.Type, y = Net.Sales, fill = Agency.Type)) +
  geom_boxplot(alpha = 0.7, outlier.alpha = 0.3) +
  stat_summary(fun = mean, geom = "point", shape = 18, size = 3, color = "red") +
  labs(title = "Net Sales by Agency Type",
       subtitle = "Red diamonds show group means",
       x = "Agency Type",
       y = "Net Sales",
       fill = "Agency Type") +
  theme_custom +
  scale_fill_brewer(palette = "Set2")

# Boxplot for Distribution Channel 

p2 <- ggplot(travel_data, aes(x = Distribution.Channel, y = Net.Sales, fill = Distribution.Channel)) +
  geom_boxplot(alpha = 0.7, outlier.alpha = 0.3) +
  stat_summary(fun = mean, geom = "point", shape = 18, size = 3, color = "red") +
  annotate("text", x = 1, y = max(travel_data$Net.Sales), 
           label = paste("n =", channel_dist["Offline"]), vjust = -0.5, fontface = "bold") +
  annotate("text", x = 2, y = max(travel_data$Net.Sales), 
           label = paste("n =", channel_dist["Online"]), vjust = -0.5, fontface = "bold") +
  labs(title = "Net Sales by Distribution Channel",
       subtitle = "Note: Severe sample size imbalance",
       x = "Distribution Channel",
       y = "Net Sales",
       fill = "Distribution Channel") +
  theme_custom +
  scale_fill_brewer(palette = "Set1")

# Interaction plot

p3 <- ggplot(travel_data, aes(x = Agency.Type, y = Net.Sales, fill = Distribution.Channel)) +
  geom_boxplot(alpha = 0.7, outlier.alpha = 0.3) +
  labs(title = "Net Sales by Agency Type and Distribution Channel",
       subtitle = "No significant interaction effect observed",
       x = "Agency Type",
       y = "Net Sales",
       fill = "Distribution Channel") +
  theme_custom +
  scale_fill_brewer(palette = "Set1")

# Display plots

print(p1)
print(p2)
print(p3)
```
<hr style="border: 1px solid #000000;">

### **One/Two Way ANOVA Review and Selection**

The assignment requires us to:

- Perform one-way ANOVA to examine the effect of individual factors on Net Sales

- Perform two-way ANOVA to examine the combined effects of both factors

- Use AIC to determine which model provides a better fit for the dataset

- Create appropriate visualizations to support our findings

We'll analyze how Agency Type and Distribution Channel affect Net Sales in the 
travel insurance business by completing a one-way ANOVA individually, then a 
two-way ANOVA using both, and then a comparison to understand which model is best.
<hr style="border: 1px solid #000000;">

### **One-way ANOVA Analysis - Agency Type, and its Effect on Net Sales**

```{r one-way-AA-AgencyType}

# One-way ANOVA for Agency Type

one_way_agency <- aov(Net.Sales ~ Agency.Type, data = travel_data)

# Summary of the ANOVA

agency_summary <- summary(one_way_agency)
agency_summary_df <- tidy(one_way_agency)

# Display results

cat('One-Way ANOVA - Agency Type vs Net Sales:')
print(agency_summary)

# Calculate group means

agency_means <- travel_data %>%
  group_by(Agency.Type) %>%
  summarise(Mean_Net_Sales = mean(Net.Sales),
            SD_Net_Sales = sd(Net.Sales),
            Count = n()) %>%
  arrange(desc(Mean_Net_Sales))

# Display formatted group means

agency_means %>%
  mutate(across(c(Mean_Net_Sales, SD_Net_Sales), ~ paste0('$', round(., 2)))) %>%
  kable(caption = 'Agency Type Group Means') %>%
  kable_styling(bootstrap_options = c('striped', 'hover', 'condensed'), full_width = FALSE)

# Calculate effect size

ss_between <- agency_summary[[1]]$'Sum Sq'[1]
ss_total <- sum(agency_summary[[1]]$'Sum Sq')
eta_squared_agency <- ss_between / ss_total

# Assign the dominant effect size for the final summary table

eta_squared <- eta_squared_agency
```
<hr style="border: 1px solid #000000;">

#### **Summary**

The one-way ANOVA for Agency Type shows a highly significant effect on Net Sales
(F = `r round(agency_summary[[1]]$'F value'[1], 2)`,
p < `r ifelse(agency_summary[[1]]$'Pr(>F)'[1] < 0.001, "0.001", round(agency_summary[[1]]$'Pr(>F)'[1], 3))`).
Airlines have substantially higher average Net Sales (`r paste0('$', round(agency_means$Mean_Net_Sales[agency_means$Agency.Type == 'Airlines'], 2))`) compared to Travel Agencies (`r paste0('$', round(agency_means$Mean_Net_Sales[agency_means$Agency.Type == 'Travel Agency'], 2))`).

The effect size ($\eta^2$ = `r round(eta_squared_agency, 4)`) indicates that Agency Type explains `r round(eta_squared_agency * 100, 2)`% of the variance in Net Sales.
<hr style="border: 1px solid #000000;">

### **One-way ANOVA Analysis - Distribution Channel, and its Effect on Net Sales**

```{r one-way-AA-DistChannel}

# One-way ANOVA for Distribution Channel

one_way_channel <- aov(Net.Sales ~ Distribution.Channel, data = travel_data)

# Summary of the ANOVA

channel_summary <- summary(one_way_channel)
channel_summary_df <- tidy(one_way_channel)

# Display results

cat('One-Way ANOVA - Distribution Channel vs Net Sales:')
print(channel_summary)

# Calculate group means

channel_means <- travel_data %>%
  group_by(Distribution.Channel) %>%
  summarise(Mean_Net_Sales = mean(Net.Sales),
            SD_Net_Sales = sd(Net.Sales),
            Count = n()) %>%
  arrange(desc(Mean_Net_Sales))

# Display formatted group means
channel_means %>%
  mutate(across(c(Mean_Net_Sales, SD_Net_Sales), ~ paste0('$', round(., 2)))) %>%
  kable(caption = 'Distribution Channel Group Means') %>%
  kable_styling(bootstrap_options = c('striped', 'hover', 'condensed'), full_width = FALSE)

# Calculate effect size

ss_between <- channel_summary[[1]]$'Sum Sq'[1]
ss_total <- sum(channel_summary[[1]]$'Sum Sq')
eta_squared_channel <- ss_between / ss_total
```
<hr style="border: 1px solid #000000;">

#### **Summary**

The one-way ANOVA for Distribution Channel also shows a statistically significant effect on Net Sales
(F = `r round(channel_summary[[1]]$'F value'[1], 2)`,
p `r if(channel_summary[[1]]$'Pr(>F)'[1] < 0.001) {"< 0.001"} else {paste("=", round(channel_summary[[1]]$'Pr(>F)'[1], 3))}`).
Online channels have a higher average Net Sales (`r paste0('$', round(channel_means$Mean_Net_Sales[channel_means$Distribution.Channel == 'Online'], 2))`) compared to Offline channels (`r paste0('$', round(channel_means$Mean_Net_Sales[channel_means$Distribution.Channel == 'Offline'], 2))`).

However, the effect size (η² = `r round(eta_squared_channel, 4)`) is smaller than Agency Type, indicating that Distribution Channel explains only `r round(eta_squared_channel * 100, 2)`% of the variance in Net Sales.
<hr style="border: 1px solid #000000;">

### **Model Comparison**

To determine the best model, we compare four competing models: both one-way models,
a two-way additive model, and a two-way interaction model, using the Akaike 
Information Criterion (AIC). The lowest AIC value indicates the preferred model 
due to its better balance of fit and complexity.

```{r model-comparison}

# One-Way ANOVA Models (Both Factors)

model_oneway_agency <- aov(Net.Sales ~ Agency.Type, data = travel_data)
model_oneway_channel <- aov(Net.Sales ~ Distribution.Channel, data = travel_data)

# Two-Way ANOVA (Additive/Main Effects Only)

model_twoway_additive <- aov(Net.Sales ~ Agency.Type + Distribution.Channel, data = travel_data)

# Two-Way ANOVA (With Interaction) 

model_twoway_interaction <- aov(Net.Sales ~ Agency.Type * Distribution.Channel, data = travel_data)

# Calculate AIC for all models

aic_oneway_agency <- AIC(model_oneway_agency)
aic_oneway_channel <- AIC(model_oneway_channel)
aic_twoway_additive <- AIC(model_twoway_additive)
aic_twoway_interaction <- AIC(model_twoway_interaction)

# Function to calculate partial eta squared for ANOVA models

calculate_partial_eta_squared <- function(anova_model) {
  anova_summary <- summary(anova_model)
  if (length(anova_summary) == 1) {
    ss_effect <- anova_summary[[1]]$'Sum Sq'[1]
    ss_total <- sum(anova_summary[[1]]$'Sum Sq')
    return(ss_effect / ss_total)
  } else {
    ss_effects <- anova_summary[[1]]$'Sum Sq'[1:(nrow(anova_summary[[1]])-1)]
    ss_residual <- anova_summary[[1]]$'Sum Sq'[nrow(anova_summary[[1]])]
    partial_eta_squared <- ss_effects / (ss_effects + ss_residual)
    return(partial_eta_squared)
  }
}

# Calculate effect sizes for all models

eta_agency <- calculate_partial_eta_squared(model_oneway_agency)
eta_channel <- calculate_partial_eta_squared(model_oneway_channel)

# For two-way models, we'll use the Agency Type effect size for comparison

eta_additive <- calculate_partial_eta_squared(model_twoway_additive)
eta_interaction <- calculate_partial_eta_squared(model_twoway_interaction)

# Extract just the Agency Type effect size from two-way models for consistent comparison

eta_additive_agency <- if(length(eta_additive) > 1) eta_additive[1] else eta_additive
eta_interaction_agency <- if(length(eta_interaction) > 1) eta_interaction[1] else eta_interaction

# Create a comprehensive comparison table

aic_comparison <- data.frame(
  Model = c('One-Way (Agency Type)', 'One-Way (Distribution Channel)', 'Two-Way (Additive)', 'Two-Way (Interaction)'),
  Formula = c('Net.Sales ~ Agency.Type', 'Net.Sales ~ Distribution.Channel', 
              'Net.Sales ~ Agency.Type + Distribution.Channel', 
              'Net.Sales ~ Agency.Type * Distribution.Channel'),
  AIC = c(aic_oneway_agency, aic_oneway_channel, aic_twoway_additive, aic_twoway_interaction),
  Effect_Size = c(
    round(eta_agency, 4),
    round(eta_channel, 4),
    round(eta_additive_agency, 4),
    round(eta_interaction_agency, 4)
  )
)

# Identify the best model based on the lowest AIC

best_model_name <- aic_comparison$Model[which.min(aic_comparison$AIC)]
best_aic <- round(min(aic_comparison$AIC), 2)

# Print the name of the best model for the final summary table

best_model <- best_model_name

# Display the AIC Comparison Table

cat('AIC Comparison of ANOVA Models')
aic_comparison %>%
  mutate(
    AIC = round(AIC, 2),
    Effect_Size = round(Effect_Size, 4)
  ) %>%
  arrange(AIC) %>%
  kable(caption = 'Model Selection using AIC (Lower is Better)') %>%
  kable_styling(bootstrap_options = c('striped', 'hover', 'condensed'), full_width = FALSE) %>%
  row_spec(which(aic_comparison$Model == best_model_name), bold = TRUE, color = 'white', background = 'green')

# Print the summary of the best model

if (best_model_name == 'One-Way (Agency Type)') {
  best_model_aov <- model_oneway_agency
} else if (best_model_name == 'One-Way (Distribution Channel)') {
  best_model_aov <- model_oneway_channel
} else if (best_model_name == 'Two-Way (Additive)') {
  best_model_aov <- model_twoway_additive
} else {
  best_model_aov <- model_twoway_interaction
}

cat('ANOVA Summary for the AIC-Selected Best Model: ', best_model_name, '')
print(summary(best_model_aov))

# Create AIC comparison visualization

aic_plot_data <- aic_comparison %>%
  mutate(
    Model = factor(Model, levels = Model[order(AIC)]),
    AIC_Difference = AIC - min(AIC)
  )

# Create Plots

ggplot(aic_plot_data, aes(x = reorder(Model, AIC), y = AIC, color = Model == best_model_name)) +
  geom_point(size = 4) +
  geom_segment(aes(xend = reorder(Model, AIC), yend = min(AIC) - 100), 
               linetype = "dashed", alpha = 0.3) +
  geom_text(aes(label = paste0("AIC = ", round(AIC, 1), 
                              ifelse(Model == best_model_name, " (BEST)", ""))), 
            vjust = -0.8, hjust = 0.5, size = 3.5, fontface = "bold") +
  scale_color_manual(values = c('TRUE' = '#2E86AB', 'FALSE' = '#A23B72'), guide = 'none') +
  labs(title = 'AIC Model Comparison',
       subtitle = 'Lower AIC values indicate better model fit\nPoints show actual AIC values for precise comparison',
       x = 'Model',
       y = 'AIC Value') +
  theme_custom +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  ylim(min(aic_plot_data$AIC) - 100, max(aic_plot_data$AIC) + 100)

# Create Simple Table

aic_ranked <- aic_comparison %>%
  arrange(AIC) %>%
  mutate(
    Rank = 1:n(),
    Delta_AIC = AIC - min(AIC),
    AIC = round(AIC, 1),
    Delta_AIC = round(Delta_AIC, 1)
  )

cat('AIC Model Ranking (Best to Worst)')
kable(aic_ranked[, c('Rank', 'Model', 'AIC', 'Delta_AIC')], 
      caption = 'Model comparison using Akaike Information Criterion') %>%
  kable_styling(bootstrap_options = c('striped', 'hover'), full_width = FALSE) %>%
  row_spec(1, bold = TRUE, color = 'white', background = '#2E86AB')

# Interpretation

cat('Model Selection Interpretation')
cat('The AIC comparison reveals that', best_model_name, 'provides the best fit with AIC =', best_aic, '.\n')
cat('The Agency Type effect size remains consistently strong across all models (η² ≈', 
    round(mean(c(eta_agency, eta_additive_agency, eta_interaction_agency)), 3),'), confirming its dominance.\n')

# Assign the dominant effect size for the final summary table

eta_squared <- eta_agency
```
<hr style="border: 1px solid #000000;">

#### **Summary**

The AIC comparison reveals that the `r best_model_name` provides the best balance
of model fit and complexity. Notably:

- **Agency Type one-way ANOVA** (AIC = `r round(aic_oneway_agency, 2)`) explains
`r round(eta_squared_agency * 100, 2)`% of variance

- **Distribution Channel one-way ANOVA** (AIC = `r round(aic_oneway_channel, 2)`)
explains only `r round(eta_squared_channel * 100, 2)`% of variance

- The `r ifelse(best_model_name == "One-Way (Agency Type)", "simpler **one-way Agency Type model** outperforms more complex models", "more complex model provides the best fit")`

This confirms that Agency Type is not only statistically significant but also the
most practically important factor for understanding Net Sales variation.
<hr style="border: 1px solid #000000;">

### **Assumption Check**

```{r check-assumptions}

# Use one way Agency Type as the selected model

residuals_data <- residuals(one_way_agency)

# Plots

p4 <- ggplot(data.frame(Residuals = residuals_data), aes(x = Residuals)) +
  geom_histogram(bins = 50, fill = "steelblue", alpha = 0.7) +
  labs(title = "Distribution of Residuals",
       subtitle = "Check for approximate normality",
       x = "Residuals",
       y = "Frequency") +
  theme_custom

# QQ Plot

p5 <- ggplot(data.frame(Residuals = residuals_data)) +
  stat_qq(aes(sample = Residuals), alpha = 0.6, size = 0.5) +
  stat_qq_line(aes(sample = Residuals), color = "red", size = 1) +
  labs(title = "Q-Q Plot of Residuals",
       subtitle = "Comparison of observed residuals vs theoretical normal distribution",
       x = "Theoretical Quantiles",
       y = "Sample Quantiles") +
  theme_custom

# Display plots separately to prevent crowding

print(p4)
print(p5)

# Variance comparison

variance_check <- travel_data %>%
  group_by(Agency.Type) %>%
  summarise(
    N = n(),
    Variance = var(Net.Sales),
    SD = sd(Net.Sales),
    .groups = 'drop'
  )

# Calculate variance ratio

variance_ratio <- variance_check$Variance[variance_check$Agency.Type == "Airlines"] /
                  variance_check$Variance[variance_check$Agency.Type == "Travel Agency"]

cat('Variance Comparison by Agency Type:')
print(variance_check)
cat('Variance Ratio (Airlines/Travel Agency):', round(variance_ratio, 2), '')

# Variance Assessment

if(variance_ratio < 2) {
  variance_assessment <- "Excellent (ratio < 2)"
} else if(variance_ratio < 4) {
  variance_assessment <- "Acceptable (ratio 2-4)"
} else {
  variance_assessment <- "Potential concern (ratio ≥ 4)"
}

cat('Variance Homogeneity Assessment:', variance_assessment, '')

# Residual statistics

cat('Residual Summary Statistics:')
residual_summary <- summary(residuals_data)
print(residual_summary)

# Check if residuals are centered around zero

cat('Mean of residuals:', round(mean(residuals_data), 6), '(should be approximately 0)')

# Check residual skewness and kurtosis with interpretation

residual_skewness <- moments::skewness(residuals_data)
residual_kurtosis <- moments::kurtosis(residuals_data)

cat('Residual skewness:', round(residual_skewness, 3), 
    ifelse(abs(residual_skewness) > 2, '***SUBSTANTIAL DEVIATION FROM NORMALITY***', 
           ifelse(abs(residual_skewness) > 1, '(moderate deviation)', '(near normal)')), '\n')

cat('Residual kurtosis:', round(residual_kurtosis, 3), 
    ifelse(abs(residual_kurtosis - 3) > 2, '***SUBSTANTIAL DEVIATION FROM NORMALITY***', 
           ifelse(abs(residual_kurtosis - 3) > 1, '(moderate deviation)', '(near normal)')), '\n')

# Formal assessment

cat('\n--- FORMAL ASSUMPTION ASSESSMENT ---\n')
concerns <- c()
if(variance_ratio >= 4) concerns <- c(concerns, "Heteroscedasticity (variance ratio ≥ 4)")
if(abs(residual_skewness) > 2) concerns <- c(concerns, "Substantial skewness")
if(abs(residual_kurtosis - 3) > 2) concerns <- c(concerns, "Substantial kurtosis")

if(length(concerns) == 0) {
  cat('✓ All assumptions reasonably met for large sample ANOVA\n')
} else {
  cat('⚠️ Potential concerns:\n')
  for(concern in concerns) cat('  -', concern, '\n')
  cat('\nNote: Large sample size (N =', nrow(travel_data), ') provides some robustness to these violations.\n')
}
```
<hr style="border: 1px solid #000000;">

#### **Summary**

The diagnostic assessment reveals:

- **Sample Size**: Large dataset (N = `r nrow(travel_data)`) provides substantial robustness to assumption violations

- **Normality**: Visual inspection shows some deviation from perfect normality in the Q-Q plot

- **Variance Comparison**: Airlines show `r round(variance_ratio, 1)`x the variance of Travel Agencies

- **Variance Assessment**: `r variance_assessment`

<hr style="border: 1px dashed #000000;">

**Key Findings Requiring Attention:**

- **Skewness**: `r round(residual_skewness, 3)` (substantial deviation from 0) ⚠️
- **Kurtosis**: `r round(residual_kurtosis, 3)` (substantial deviation from 3) ⚠️  
- **Variance Ratio**: `r round(variance_ratio, 1)` (exceeds recommended threshold of 4) ⚠️

<hr style="border: 1px dashed #000000;">

**Interpretation & Mitigation:**

While the large sample size provides some protection against these assumption violations, the substantial skewness, kurtosis, and heteroscedasticity indicate the results should be interpreted with caution. Though I did not do any data cleaning due to the protection that the large data set provided. I did consider the following before making that decision:

- Data transformations (log, square root) to address skewness and variance issues
- Non-parametric alternatives (Kruskal-Wallis test)
- Bootstrapping methods for confidence intervals

The current ANOVA results remain informative but represent approximations rather
than exact inferences given these assumption violations.

### **Final Conclusions and Recommendations**

```{r summary}

# Calculate performance difference

performance_diff_value <- round(agency_means$Mean_Net_Sales[agency_means$Agency.Type == 'Airlines'] - agency_means$Mean_Net_Sales[agency_means$Agency.Type == 'Travel Agency'], 2)

# Create comprehensive summary

final_summary <- data.frame(
  Aspect = c(
    'Primary Finding',
    'Statistical Significance',
    'Effect Size (η²)',
    'Mean Net Sales - Airlines',
    'Mean Net Sales - Travel Agencies', 
    'Performance Difference',
    'Best Model (AIC)',
    'AIC Model Ranking',
    'Data Quality Note',
    'Strategic Recommendation'
  ),
  Result = c(
    'Agency Type is the dominant factor',
    'Highly significant (p < 0.001)',
    paste(round(eta_squared * 100, 1), '% of variance explained'),
    paste0('$', round(agency_means$Mean_Net_Sales[agency_means$Agency.Type == 'Airlines'], 2)),
    paste0('$', round(agency_means$Mean_Net_Sales[agency_means$Agency.Type == 'Travel Agency'], 2)),
    paste0('$', performance_diff_value, ' higher for Airlines'),
    best_model,
    'Agency Type > Distribution Channel (lower AIC = better fit)',
    'Severe imbalance in Distribution Channel data',
    'Focus on Agency Type for business strategy'
  )
)

kable(final_summary, caption = 'Comprehensive Analysis Summary') %>%
  kable_styling(bootstrap_options = 'striped', full_width = FALSE) %>%
  row_spec(1, bold = TRUE, color = 'white', background = '#2E86AB') %>%
  row_spec(10, bold = TRUE, color = 'white', background = '#A23B72')
```
<hr style="border: 1px solid #000000;">

#### **Final Summary**

Based on the comprehensive ANOVA analysis:

- Agency Type is the most important factor affecting Net Sales, explaining `r round(eta_squared * 100, 1)`%
of the variance

- Airlines significantly outperform Travel Agencies in Net Sales by `r paste0('$', round(agency_means$Mean_Net_Sales[agency_means$Agency.Type == "Airlines"] - agency_means$Mean_Net_Sales[agency_means$Agency.Type == "Travel Agency"], 2))`
on average

- Distribution Channel shows statistical significance but results are unreliable
due to severe data imbalance (98.2% Online vs 1.8% Offline)

- No interaction effect exists between Agency Type and Distribution Channel

- The one-way Agency Type model provides the most parsimonious and reliable explanation of Net Sales variation

<hr style="border: 3px solid #000000;">

## **Reflection**

### **Key Learning Takeaways**:

This week solidified that ANOVA is a powerful extension of t-tests for comparing multiple groups. The hands-on experience revealed that proper implementation requires attention to both statistical assumptions and technical execution in R. Most importantly, I learned that model comparison approaches like AIC provide valuable insights beyond individual hypothesis tests, and that clear communication of methodological choices is essential for reproducible analysis.
<hr style="border: 1px solid #CCCCCC;">

#### **Conceptually Intuitive/Easier Aspects**

This week's ANOVA coursework contained several elements that felt conceptually straightforward:

- The fundamental logic of ANOVA - comparing between-group variance to within-group variance to detect significant differences - was intuitive and built nicely on previous t-test concepts.

- Setting up one-way ANOVA models in R felt natural, as the formula syntax (Net.Sales ~ Agency.Type) clearly expressed the relationship we were testing.

- Interpreting ANOVA summary outputs was relatively straightforward, with F-statistics and p-values following the same interpretation framework I've learned in past courses and used throughout this course.

- Calculating basic effect sizes (η²) from the sum of squares made conceptual sense as a measure of practical importance beyond statistical significance.

<hr style="border: 1px solid #CCCCCC;">

#### **More Challenging Aspects**

Several aspects required deeper statistical understanding and technical problem-solving:

- Understanding the differences between one-way and two-way ANOVA conceptually, particularly how two-way ANOVA handles multiple factors and tests for interaction effects.

- Implementing proper model comparison using AIC required careful attention to ensure all models were correctly specified and comparable.

- Diagnostic checking and assumption verification proved complex, particularly interpreting Q-Q plots, understanding variance homogeneity requirements, and determining the practical impact of violations in large samples.

- Debugging R Markdown technical issues consumed significant time, especially troubleshooting inline coding LaTeX notation for Greek letters, then HTML coding for superscript, finally resulting in me copying and pasting Unicode symbols to create $\eta^2$ for one of my data summaries.

- Creating effective visualizations that clearly communicated the ANOVA results and model comparisons required experimentation with different ggplot approaches, and a reminder sometimes a simple table comparison is best suited.

<hr style="border: 1px solid #CCCCCC;">

#### **Start/Stop/Continue**

**START**:

- Practicing with different types of post-hoc tests to understand when they're needed

- Exploring additional diagnostic plots and tests for ANOVA assumptions

- Building a personal code repository for common workflows

**STOP**:

- Overcomplicating interpretations with unnecessary business context when the focus is methodological

- Using complex notation when simpler alternatives achieve the same pedagogical purpose

- Spending excessive time on formatting issues before ensuring the statistical content is correct

**CONTINUE**:

- The systematic approach to checking assumptions before interpreting results

- Comparing multiple models to understand which provides the best fit

- Using both statistical outputs and visualizations to tell a complete analytical story

- Documenting the decision-making process throughout the analysis

<hr style="border: 3px solid #000000;">